---
title: "Introduction to ggplot"
output: html_notebook
---

## Plotting with ggplot

Layers:

1. Data: The dataset being summarised
2. Geometry: The type of plot
3. Aesthetic mapping: Variables mapped to visual cues, such as x/y-axis values and colour

Others:

- Scale
- Labels, Title, Legend
- Themes/Styles

### Basics

```{r}
library(tidyverse)
data(murders)
```

Without describing anything, `ggplot` produces a blank slate

```{r}
murders %>% ggplot()
```
Printing the `ggplot` object gives the same output

```{r}
p <- ggplot(data = murders)
class(p)
print(p)
```
### Layers

Here we add a geometry layer, assigning column names to x and y

```{r}
murders %>% ggplot() +
  geom_point(aes(x = population/10^6, y = total))
```
The same effect can be achieved by using the plot object that we had created, `p`.

```{r}
p + geom_point(aes(x = population/10^6, y = total))
```
... and we can add more layers

```{r}
p + geom_point(aes(population/10^6, total)) +
    geom_text(aes(population/10^6, total, label = abb))
```
Tinkering with the layers byt passing further arguments, outside of `aes()`

```{r}
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
```
The code can be simplified by setting a global mapping on `p`. Note that local mappings override default or global mappings.

```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))

p + 
  geom_point(size = 3) +
  geom_text(nudge_x = 1.5)
```

#### Scales

Scales are another layer. Here we use the function `scale_x_continuous()` or `scale_y_continuous()` which are generic functions, but `scale_x_log10()` and `scale_y_log10()` can be used to simplify here.

```{r}
p + geom_point(size = 3) +
  geom_text(nudge_x = 0.075) + # this value must be reduced for the log plot
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")
```

#### Labels & title

The functions are `xlab` and `ylab`

```{r}
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
```

#### Colours

Redefining `p` to include everything except the points layer. Then we can define the points layer, with a colour mapping, according to a categorical variable. The `x` and `y` mappings are already defined in the function call to `ggplot()` here.

```{r}
# redefine p to be everything except the points layer
p <- murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")

# make all points blue
p + geom_point(size = 3, color = "blue")

# color points by region
p + geom_point(aes(col = region), size = 3)
```

#### Lines

If we want to combine data on a graph, such as adding a line, it can simply be added as a new layer. Here the average murder rate is calculated from `data(murders)`, and the function `geom_abline()` describes the layer. Note that for the second plot where a dashed grey line is use, the line is added before the data, as the ordering defines the visual overlapping of layers on the plot.

```{r}
# define average murder rate
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    pull(rate)
    
# basic line with average murder rate for the country
p + geom_point(aes(col = region), size = 3) +
    geom_abline(intercept = log10(r)) +    # slope is default of 1
    scale_color_discrete(name = "Region") # capitalise legend title
    
# change line to dashed and dark grey, line under points
p + 
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3) + 
    scale_color_discrete(name = "Region")
```

##### Legend title

In the previous plots, note that we added this layer to our plot object to capitalise the legend title.

```scale_color_discrete(name = "Region")    # capitalize legend title```

## Add-on Packages

Add-on packages provide additional features that fit well with the grammar in `ggplot`. Here they are simply added

```{r}
library(dslabs)
library(ggthemes)
library(ggrepel)
ds_theme_set()

# define the intercept
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    .$rate
    
# make the plot, combining all elements
p <- murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3) +
    geom_text_repel() +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010") +
    scale_color_discrete(name = "Region")

# themes from ggthemes
p + theme_economist()    # style of the Economist magazine
p + theme_fivethirtyeight()    # style of the FiveThirtyEight website
```

## Examples based on Distributions notebook (`01.md`)

```{r}
# load heights data
library(tidyverse)
library(dslabs)
data(heights)

# define p
p <- heights %>%
    filter(sex == "Male") %>%
    ggplot(aes(x = height))
    
# basic histograms
p + geom_histogram()
p + geom_histogram(binwidth = 1)

# histogram with blue fill, black outline, labels and title
p + geom_histogram(binwidth = 1, fill = "blue", col = "black") +
    xlab("Male heights in inches") +
    ggtitle("Histogram")
```

