---
title: "R Notebook"
output: html_notebook
---

## Summarising with `dplyr`

`dplyr` is a package for dataframe manipulation.

### `summarize`

`summarize()` computes summary statistics from the data frame. It returns a data frame whose column names are defined within the function call.

```{r}
library(tidyverse)
library(dslabs)
data(heights)

# compute average and standard deviation for males
s <- heights %>%
    filter(sex == "Male") %>%
    summarize(average = mean(height), standard_deviation = sd(height))
    
# access average and standard deviation from summary table
s$average
s$standard_deviation

# compute median, min and max
heights %>%
    filter(sex == "Male") %>%
    summarize(median = median(height),
              minimum = min(height),
              maximum = max(height))
# alternative way to get min, median, max in base R
quantile(heights$height, c(0, 0.5, 1))

# NOTE: The following code will NOT generate an error if using dplyr 1.0 or later

# generates an error: summarize can only take functions that return a single value
heights %>%
    filter(sex == "Male") %>%
    summarize(range = quantile(height, c(0, 0.5, 1)))
```

### The Dot Placeholder

The dot operator allows you to access values stored in data that is being piped in using the %>% character. The dot is a placeholder for the data being passed in through the pipe.

Note that an equivalent way to extract a single column using the pipe is `us_murder_rate %>% pull(rate)`. The `pull()` function will be used in later course material.

```{r}
library(tidyverse)
library(dslabs)
data(murders)

murders <- murders %>% mutate(murder_rate = total/population*100000)
summarize(murders, mean(murder_rate))

# calculate US murder rate, generating a data frame
us_murder_rate <- murders %>%
    summarize(rate = sum(total) / sum(population) * 100000)
    us_murder_rate

# extract the numeric US murder rate with the dot operator
us_murder_rate %>% .$rate

# calculate and extract the murder rate with one pipe
us_murder_rate <- murders %>%
    summarize(rate = sum(total) / sum(population) * 100000) %>%
    .$rate
```

## Group By

  The group_by() function from dplyr  converts a data frame to a grouped data frame, creating groups using one or more variables.

  summarize() and some other dplyr functions will behave differently on grouped data frames.
  
  Using summarize() on a grouped data frame computes the summary statistics for each of the separate groups.

```{r}
# libraries and data
library(tidyverse)
library(dslabs)
data(heights)
data(murders)

# compute separate average and standard deviation for male/female heights
heights %>%
    group_by(sex) %>%
    summarize(average = mean(height), standard_deviation = sd(height))

# compute median murder rate in 4 regions of country
murders <- murders %>%
    mutate(murder_rate = total/population * 100000)
murders %>%
    group_by(region) %>%
    summarize(median_rate = median(murder_rate))
```

## Sorting Data Tables

The `arrange()` function from dplyr sorts a data frame by a given column.

You can `arrange()` by multiple levels: within equivalent values of the first level, observations are sorted by the second level, and so on.

The `slice_min()`/`slice_max()` functions shows the top/bot results ranked by a given variable, but the results are not ordered. You can combine them with `arrange()` to return the top results in order.

```{r}
# libraries and data
library(tidyverse)
library(dslabs)
data(murders)

# set up murders object
murders <- murders %>%
    mutate(murder_rate = total/population * 100000)
    
# arrange by population column, smallest to largest
murders %>% arrange(population) %>% head()

# arrange by murder rate, smallest to largest
murders %>% arrange(murder_rate) %>% head()

# arrange by murder rate in descending order
murders %>% arrange(desc(murder_rate)) %>% head()

# arrange by region alphabetically, then by murder rate within each region
murders %>% arrange(region, murder_rate) %>% head()

# show the top 10 states with highest murder rate, not ordered by rate
murders %>% top_n(10, murder_rate)

# show the top 10 states with highest murder rate, ordered by rate
murders %>% arrange(desc(murder_rate)) %>% top_n(10)

# alternatively, can use the slice_max function
murders %>% slice_max(murder_rate, n = 10)
```

